---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Initialization

This step involves loading required libraries, setting the working directory, and making other
adjustments so our code will function properly:
```{r}
packages <- c("data.table", "ggplot2", "scales")
sapply(packages, require, character.only = TRUE, warn.conflicts = FALSE, quietly = TRUE)

previousWD <- getwd()
newWD <- getSrcDirectory(function(x) {x})
setwd(newWD)

options(datatable.print.nrows = 50)
```

## Loading and preprocessing the data

#### 1) Load the data

In this assignment, we use the `data.table` package, which provides high performance file loading
and methods to process tabular data. Since the data file is zipped, we first unzip it, then read the
data file, using the `fread()` function. Once the data is loaded, the CSV data file is deleted.
```{r}
dataFileName <- "activity.csv"
zipFIlename <- "activity.zip"
dataFile <- unzip(zipFIlename)
data <- fread(dataFileName)
unlink(dataFile)
print(data)
```

#### 2) Process/transform the data (if necessary) into a format suitable for your analysis

At this point, no further processing is required.

## What is mean total number of steps taken per day?

#### 1) Calculate the total number of steps taken per day

This can be accomplished in a single line of code using the built-in `data.table` syntax.

```{r}
stepsPerDay <- data[, .(total = sum(steps, na.rm = TRUE)), by = .(date)]
print(stepsPerDay)
```

#### 2) Make a histogram of the total number of steps taken each day

For all graphics in this report, we'll use the `ggplot2` package.

```{r, fig.align='center'}
g <- ggplot(stepsPerDay, aes(total)) + 
  geom_histogram(fill = "#5588BB", color = "#336699", size = 0.8, binwidth = 1000) + 
  labs(x = "Steps Taken per Day", y = "Number of Days", title = "Histogram for Steps Taken per Day") + 
  scale_x_continuous(breaks = seq(0, 22000, 2000)) + 
  scale_y_continuous(breaks = seq(0, 10, 2))
print(g)
```

#### 3) Calculate and report the mean and median of the total number of steps taken per day

Again, the `data.table` syntax helps us calculate these results with a single line of code.
```{r}
meanMedian <- stepsPerDay[, .(mean = mean(total), median = median(total))]
print(meanMedian)
```

## What is the average daily activity pattern?

#### 1) Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

First, we must calculate the average number of steps for each 5-minute interval, across all days in the dataset.
```{r}
avgFiveMinuteInt <- data[, .(mean = mean(steps, na.rm = TRUE)), by = .(interval)]
print(avgFiveMinuteInt)
```

In order to present the plot x-axis scale in a more meaningful way (i.e. in the `hh:mm` format),
we calculate an auxiliary character vector that takes a 4-digit representation of the interval strings from the 
`avgFiveMinuteInt` table and gererates Date/Time objects from them (note the date is not important. 
We are only interested in the time, but the `timeVector` is of class `POSIXlt`, so both date and time
are shown):
```{r}
timeVector <- as.data.table(strptime(sprintf("%04d", avgFiveMinuteInt$interval), "%H%M"))
print(timeVector)
```

Finally, we can plot the data series, using the `timeVector` as labels for the x-axis, after proper
formatting with the `date_format()` function of the `scales` package:
```{r, fig.align='center'}
g <- ggplot(avgFiveMinuteInt, aes(timeVector, mean)) + 
  geom_line(size = 1) + 
  labs(x = "Time", y = "Average Steps Taken", title = "Average Steps Taken per 5-minute Intervals") + 
  scale_x_datetime(labels=date_format("%H:%M", tz = Sys.timezone()))
print(g)
```

#### 2) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxSteps <- avgFiveMinuteInt[which.max(mean), .(interval, mean)]
print(maxSteps)
```

## Imputing missing values

#### 1) Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
numNAs <- sum(is.na(data$steps))
print(numNAs)
```

#### 2) Devise a strategy for filling in all of the missing values in the dataset.

In my analysis, the imputing strategy for `NA` values consisted of using the mean value for that
particular 5-minute interval across all days. One efficient way of doing this is merging the `data` and the
`avgFiveMinuteInt` tables to create one bigger data table with separate columns for the actual
number of steps taken in each 5-minute interval and the mean for that same interval. This is done
using the `data.table` join syntax.

```{r}
setkey(data, interval)
setkey(avgFiveMinuteInt, interval)
filledData <- data[avgFiveMinuteInt]
print(filledData)
```

#### 3) Create a new dataset that is equal to the original dataset but with the missing data filled in.

Since the `NA` values a going to be filled with the mean for that 5-minute interval (a floating
point number), we must first convert the `steps` column to `double`:

```{r}
filledData$steps <- as.double(filledData$steps)
```

Finally, whe filter the table to find all rows where `steps == NA` and copy the value of the `mean`
column to the `steps` column:
```{r}
filledData[is.na(steps), steps := mean]
```

To make the final data a little bit cleaner, we then remove the `mean` column and sort the new table
across the `date` column:

```{r}
filledData <- filledData[, .(steps, date, interval)]
setkey(filledData, date)
print(filledData)
```

We can see the new table has no remaining `NA` values:
```{r}
sum(is.na(filledData))
```

#### 4) Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

In order to differentiate from the preivous calculated values, we'll add a common suffix to the values
calculated in this step. The calculation process is otherwise the same as done before, but using the
`filledDate` vector instead.

```{r, fig.align='center'}
stepsPerDay2 <- filledData[, .(total = sum(steps, na.rm = T)), by = .(date)]
print(stepsPerDay2)

g <- ggplot(stepsPerDay2, aes(total)) + 
  geom_histogram(fill = "#5588BB", color = "#336699", size = 0.8, binwidth = 1000) + 
  labs(x = "Steps Taken per Day", y = "Number of Days", title = "Histogram for Steps Taken per Day") + 
  scale_x_continuous(breaks = seq(0, 22000, 2000)) + 
  scale_y_continuous(breaks = seq(0, 18, 2))
print(g)

meanMedian2 <- stepsPerDay2[, .(mean = mean(total), median = median(total))]
print(meanMedian2)
```

As we can see, the values do differ from the preivously calculated ones. The immediate effect of 
imputing the missing values is to make the mean and median the same.

## Are there differences in activity patterns between weekdays and weekends?

#### 1) Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

Here again, the `data.table` built-in functions help us adding columns to the dataset with the 
factor values. We'll use the `filledData` table for the calculations below:

```{r}
Sys.setlocale(locale = "en_US")
filledData[, c("day_of_week", "weekday") := list(weekdays(strptime(date, "%Y-%m-%d")), "Weekday")]
filledData[(day_of_week == "Saturday") | (day_of_week == "Sunday"), weekday := "Weekend"]
print(filledData)
```

#### 2) Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

First, we must calculate the new means for every 5-minute interval, using the imputed data and
grouped by the `weekday` column.

```{r}
avgFiveMinuteInt2 <- filledData[, .(mean = mean(steps, na.rm = T)), by = .(interval, weekday)]
print(avgFiveMinuteInt2)
```

Now, we use the same strategy as before to calculate a helper vector of Date/Time values to be used
as x-axis labels:
```{r}
timeVector2 <- as.data.table(strptime(sprintf("%04d", avgFiveMinuteInt2$interval), "%H%M"))
print(timeVector2)
```

Finally, we create the panel plot as explained before:
```{r, fig.align='center'}
g <- ggplot(avgFiveMinuteInt2, aes(timeVector2, mean)) + 
  geom_line(size = 1) + 
  labs(x = "Time", y = "Average Steps Taken", 
       title = "Average Steps Taken per 5-minute Intervals") +
  facet_grid(weekday ~ .) + 
  scale_x_datetime(labels=date_format("%H:%M", tz = Sys.timezone()))
print(g)
```

Finally, now that all processing is done, we set the working directory back to what it was at the
beginning of the analysis.
```{r}
setwd(previousWD)
```

